\chapter{Methods}
\section{Data Collection}
The \ac{MRO}
\section{Definition of Scope}
\subsection{Lava Flow Mapping}
The global \ac{CTX} mosaic was used to identify lava flows near the summit of \ac{OM}. Flows were outlined and mapped as polygons in ArcGIS Pro. Polygons extended only over areas where flow edges are clearly visible, since the direction of flow at each location is the key information to be derived. The ``Polygon to Centerline'' tool was used to derive linear flow paths from the polygonal flow outlines.
\subsection{Point Grid}
The ``Generate Tesselation'' tool was used to construct a grid of squares (polygon features) with area (FIX) \qty{.36}{\km\squared} over the extent of the lava flows. Of those squares whose centers lie within a flow polygon, a point feature was created using the ``Polygon to Point'' tool. The overall effect of this procedure is to generate a square grid of points that spans the entire area covered by the mapped lava flows. To each of these points will be assigned several varaibles representing various descriptive parameters of the modern and paleo surface.
\section{Surface Parameters}
I run the ``Surface Parameters'' tool on the \qty{200}{\m} DEM surface twice, once to produce a slope raster (in degrees) and once to produce an aspect raster (also in degrees, measured clockwise from due north). In both cases, I used a quadratic local surface type with a neighborhood radius of \qty{2}{\km}. I then used the ``Add Surface Informaton'' tool to assign these values to each feature in the point grid.
\section{Paleosurface Parameters}
\subsection{Paleoaspect}
The ``Near'' tool was used to assign the angle from the nearest flow path to each grid point. This angle is perpendicular to the paleoaspect (downhill direction) recorded by the flow. Thus, \ang{90} is added to the ``Near Angle'', and the result is stored modulo \ang{360} to avoid counting extra rotations in the final aspect direction. Since points in the grid appear on both sides of the centerline, the stored angle may be parallel or anti-parallel to the actual direction of flow. Based on visual inspection of flow morphologies, I assumed that all flows were oriented such that the paleosurface sloped downhill away from the caldera, not toward it. Thus, I used the ``Near'' tool again to determine the angle from the caldera to each grid point. If the difference between this angle and the stored aspect angle (absolute value) exceeded \ang{180}, I recalculated the aspect value by adding 180 (again modulo \ang{360}).

\subsection{Paleoslope}\label{paleoslope}

\def\figsize{3}
\begin{SCfigure}
    \begin{tikzpicture}
        \draw (0,0) ++ (210:\figsize) -- ++(30:2*\figsize);
        \draw[arrow] (0,0) -- ++(120:.5*\figsize);
        \draw[dotted] (-\figsize,0) -- (\figsize,0);
        \draw[dotted] (0,0) -- (0,\figsize);
        \draw (.8,0) arc (0:30:.8);
        \draw (0,.8) arc (90:120:.8);
        \path +(104:1.1) node {$\delta$};
        \path +(10:1.3) node {\textit{slope}};
        \draw (210:.3) -- ++(120:.3) -- ++(30:.3);
    \end{tikzpicture}
    \caption{ArcGIS Pro calculates the slope of a tilted \ac{DEM} surface, which is equal to the colatitude of the normal vector (complement of latitude). This angle is also known as ``dip.''}
\end{SCfigure}
It is not possible to assign a paleoslope directly from the lava flows like we can with the paleoaspect. Instead, I chose to place a constraint on the paleoslope by assuming the minimum possible surface attitude change between flow emplacement and the modern surface. Equivalently, since each surface attitude is equivalent to a unique unit normal vector, I sought to minimize the angular distance between the normal vectors corresponding to the modern and paleosurfaces. From the previous steps, we have a grid of point features, each with attributes including ``SLOPE'' (S), ``ASPECT'' (A), and ``PALEO\_ASPECT'' (A$'$). Note that these values are stored in degrees; corresponding lowercase letters are equivalent values converted into radians for trigonometric calculations.

The basic strategy for solving this optimization problem is as follows. First, the slope and aspect of the modern surface are plotted on the surface of a unit sphere as a colatitude and longitude, respectively. This point on the sphere corresponds to the unit vector normal to the surface. Note that the sine of a colatitude is equal to the cosine of a latitude, and vice versa; this substitution will be used in the following equations.

For the paleosurface, only the aspect is known. Instead of a unique point on the sphere, we have a line of longitude. To minimize the angular distance between the two surfaces, we are thus looking for the shortest great circle arc between the modern surface point and the paleosurface line.

To achieve this optimization and find the nearest intersection point, the sphere surface is converted to a gnomonic projection: one in which all great circles are mapped to straight lines in the cartesian plane. Then, we find the line perpendicular to the paleoaspect line which contains the modern surface point. The radial distance of this intersection point from the origin is reprojected onto a spherical colatitude, which is equivalent to the surface slope.

The equations take the following form. First, the line corresponding to the paleoaspect direction:

\begin{equation}
    y=x\cot a'. \label{line}
\end{equation}

The point corresponding to the modern surface has the formula
\begin{equation}
    [x,y]=[\tan s\sin a,\tan s\cos a].\label{point}
\end{equation}

The closest path from the point to the line is perpendicular to the line. It has point-slope form: 
\begin{equation}
    y-\tan s \cos a=-\tan a'(x-\tan s\sin a)
\end{equation}

Substituting from~\ref{line} to find the $x$-coordinate of the intersection point: 
\begin{gather}
    x\cot a'-\tan s \cos a=-\tan a'(x-\tan s\sin a)\nonumber\\
    \boxed{x=\frac{\tan s(\cos a+\tan a'\sin a)}{\cot a' + \tan a'}}
\end{gather}
And from~\ref{line}, the $y-$coordinate:
\begin{equation}
    \boxed{y=x\cot a'}
\end{equation}
Reprojecting back onto the sphere, the calculated colatitude (paleoslope):
\begin{equation}
    s'=\arctan\sqrt{x^2+y^2}
\end{equation}

\section{Surface \& Paleosurface Comparison}
\subsection{Deflection}
Before proceeding, we should check that the deflection (defined as the difference between aspect and paleoaspect) does not have a magnitude greater than \ang{90}. If it does, then this calculation has produced a nearest uphill slope, rather than downhill slope, which is clearly not the desired behavior. Therefore, at this stage any grid points with such deflections have their paleoslopes redefined to zero. Alternatively, a minimum slope required for lava flow can also be set here.
\subsection{Deformation}
From~\ref{paleoslope} we have generated two points on a spherical surface, each corresponding to a unit vector normal to a unique surface. The angular distance between these vectors defines the attribute ``DEFORM'' for each grid point. The formula is:
\begin{equation}
    \arccos\left[\cos s \cos s' + \sin s \sin s' \cos(a - a')\right] 
\end{equation}
converted into degrees for comparison with other attributes.
\subsection{Elevation Change}
\subsection{Volumetric Inflation}