\chapter{Methods}\label{cha:methods}

In this chapter, I first introduce key considerations which inform subsequent methods (Section~\ref{sec:considerations}) and define the study area (Section~\ref{sec:study-area}).

Next, I model the surface deformation signature of a magma reservoir pressure change event (Section~\ref{sec:modeling}). I pay particular attention to expressing surface displacement in terms of \emph{attitude} (orientation) change for direct comparison with topographically discordant features. The intermediate products of this section are two methods for computing tilt as a function of distance from the reservoir inflation center: an analytical solution to efficiently identify plausible inflation center locations and a numerical solution for estimating reservoir parameters like reservoir depth, size, shape, and pressure change.

Next, I construct tilt-distance datasets (Section~\ref{sec:mapping}) from discordant features in the study area to compare with model tilt solutions. I first compare the orientation of flow features (via imagery) with the surface attitude (via topography) of their surroundings. From these data I derive the first intermediate product of this section: a spatial distribution of discordant features in the study area. However, distance and tilt (for comparison with model solutions from Section~\ref{sec:modeling}) each depend on the position of the magma reservoir responsible for surface deformation. Since I do not know this position, I define an array of candidate center locations to test. For each center, I calculate a) distance from, and b) tilt implied at, each lava feature. The final intermediate product of this section is an array of tilt-distance datasets (one for each center candidate location) for analysis under the framework derived in Section~\ref{sec:modeling}.

Finally, I analyze and evaluate (Section~\ref{sec:evaluation}) each candidate center by the degree to which its associated tilt-distance dataset (Section~\ref{sec:mapping}) captures both a) the mapped discordant features and b) a plausible tilt-distance solution from Section~\ref{sec:modeling}. I use the spatial distribution of these evaluation results to assess the subsurface conditions most likely responsible for discordant features at the surface. After identifying the most likely reservoir position, I assess additional reservoir parameters most consistent with the tilt implied at the surface.

Unless otherwise stated, I perform all calculations in a Jupyter Notebook environment using a custom Python (v. 3.10.9) module \hltt{topodisc}, printed in Appendix~\ref{app:code}.

\input{chapters/methods/sec:considerations.tex}
\input{chapters/methods/sec:study-area.tex}
\input{chapters/methods/sec:modeling.tex}
\input{chapters/methods/sec:mapping.tex}
\input{chapters/methods/sec:evaluation.tex}

\section{Summary}

The final product of this section is a transformation from mapped attitude data (Figure~\ref{fig:attitude-data}) at a particular location $(\acs{lat}, \acs{lon})$ to a tilt-distance data point corresponding to a single inflation center candidate $(\acs{latC}, \acs{lonC})$. Figure~\ref{fig:tilt-example} illustrates the final results. Repeating the tilt and distance calculations for each sample (within a discordant population of interest) to produce a full tilt-distance dataset corresponding to a single center candidate. Finally, repeating this dataset construction for each center allows the full array of candidates to be evaluated and compared.